---
title: "Using ggerror"
output:
  rmarkdown::html_vignette:
    css: vignette.css
vignette: >
  %\VignetteIndexEntry{Using ggerror}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 4,
  dpi = 96,
  warning = FALSE,
  message = FALSE
)
```

```{r viz-setup, include=FALSE}
library(ggplot2)
library(ggerror)

set_theme(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(
        family = "mono",
        size = 11,
        hjust = 0)
      )
  )
```
## Why `ggerror`?

`ggplot2` ships several error geoms. `ggerror` started as a wrapper to simplify their usage, by dropping `x/y`-`min/max` with a one aesthetics **`error`**.
This doesn't just simplify the API, but it also shift the focus to the actual intention: to visualize some magnitude as error that extends from a specified value (or a central metric, more on that at `vignette("use-cases")`).

`ggerror` is no longer just a wrapper, but it offers extra functionality compared to core `ggplot2`'s error geoms.

This vignette demonstrates the basic usage of the package. For `stat_error()` summaries and `sign_aware` erros, see `vignette("use-cases")`.

## Setup

We use `airquality` — Daily air quality measurements in New York, May to September 1973.

This dataset has 153 rows with 6 columns: 4 continuous measurements,
Month and Day.

```{r data-prep}
data("airquality"); airq <- airquality

# It wouldn't be an R workflow without minimal data cleaning...
day_in_month <- function(day_in_month, month, year) {
  days_abbr <- format(as.Date(sprintf("%d-%02d-%02d", year, month, day_in_month)), "%a")
  factor(days_abbr, levels = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun"), ordered = TRUE)

}
airq$Day <- day_in_month(airq$Day, airq$Month, 1973)
airq$Month <- factor(airq$Month, labels = month.abb[5:9])
```

## Simple and defaults

A single `error` aesthetic is enough — `ggerror` infers orientation from the data (here: numeric x axis, discrete y axis) and picks `geom_errorbar` as the default base geom.


```{r simple-default}
# Subset the data to first week of May only
may_week <- subset(airq[1:7,],airq[1:7,]$Month=='May')

ggplot(may_week, aes(x = Temp, y = Day)) +
  geom_point() +
  geom_error(aes(error = sd(airq$Temp)), width = 0.2) +
  labs(title = "ggplot(airquality, aes(x = Temp, y = Day) +\ngeom_error(error = sd(Temp))",
       x=NULL, y=NULL)
```
*I did modifiy the data a little bit to make for demo porpuses*.

Same data, pinned `geom_error_pointrange()` wrapper:

```{r simple-pinned}
ggplot(co2_sum, aes(uptake.m, Type, colour = Treatment)) +
  geom_error_pointrange(aes(error = uptake.s),
                        position = position_dodge(0.4),
                        size = 0.6) +
  labs(title = "geom_error_pointrange(aes(error = uptake.s))")
```

Same data, swapped via the `error_geom` argument:

```{r simple-arg}
p +
  geom_error(aes(error = uptake.s),
             error_geom = "linerange",
             position = position_dodge(0.4),
             linewidth = 1) +
  labs(title = "geom_error(error_geom = \"linerange\")")
```

::: tip
💡 **Tip:** Because `error_geom` is a regular argument, you can iterate over it functionally — useful when scanning a few presentations to pick one:

```{r fp-tip, eval = FALSE}
purrr::map(c("errorbar", "linerange", "crossbar", "pointrange"),
           ~ p + geom_error(aes(error = uptake.s), error_geom = .x))
```
:::

## Asymmetric errors

`error_neg` and `error_pos` extend in opposite directions regardless of orientation. They're useful when the dispersion measure on each side carries different meaning — say, **standard deviation** on the low side and **standard deviation × 1.5** on the high side, or here, **SD** vs **√Var** as a sanity check that they're equal.

```{r asym-sd-vs-var}
ggplot(co2_sum, aes(uptake.m, Type)) +
  geom_point(size = 3) +
  geom_error(aes(error_neg = uptake.s,
                 error_pos = sqrt(uptake.v)),
             colour_neg = "steelblue",
             colour_pos = "firebrick",
             linewidth  = 1) +
  facet_wrap(~ Treatment) +
  labs(title = "error_neg = sd ; error_pos = sqrt(var) — by Treatment")
```

The same pattern works for **standard error vs confidence interval**, **IQR halves**, or any other paired dispersion measures — swap the two `error_*` expressions, the rest of the call stays the same.

Per-side styling extends to `colour`, `fill`, `linewidth`, `linetype`, `alpha`, and `width`. Pass either the side-specific form (`colour_neg`, `width_pos`) or the shared form (`colour`, `width`); side-specific wins.

```{r asym-crossbar}
ggplot(co2_sum, aes(uptake.m, Type)) +
  geom_error(aes(error_neg = uptake.s,
                 error_pos = sqrt(uptake.v)),
             error_geom = "crossbar",
             fill_neg = "grey85", fill_pos = "grey45",
             width_neg = 0.3,     width_pos = 0.6) +
  facet_wrap(~ Treatment) +
  labs(title = "crossbar with per-side fills + widths")
```

## One-sided bars

Suppose downstream stakeholders only care about **shortfall** — how far each group falls *below* its mean uptake — and treat overshoot as uninteresting. Set the unused side to `NA` and `ggerror` suppresses the cap and stem on that side automatically.

```{r one-sided-basic}
ggplot(co2_sum, aes(uptake.m, Type, colour = Treatment)) +
  geom_point(size = 3, position = position_dodge(0.4)) +
  geom_error(aes(error_neg = uptake.s, error_pos = NA),
             position = position_dodge(0.4),
             linewidth = 1) +
  labs(title = "downside only — error_pos = NA")
```

A linerange variant — heavier line, no caps — reads cleanly when overlaid on raw points:

```{r one-sided-linerange}
ggplot(CO2, aes(uptake, Type)) +
  geom_jitter(height = 0.2, alpha = 0.3) +
  geom_error_linerange(data = co2_sum,
                       aes(x = uptake.m,
                           error_neg = uptake.s,
                           error_pos = NA),
                       linewidth  = 1.5,
                       colour     = "firebrick") +
  facet_wrap(~ Treatment) +
  labs(title = "shortfall over raw observations")
```

A pointrange variant marks the mean and the lower bound in one go:

```{r one-sided-pointrange}
ggplot(co2_sum, aes(uptake.m, Type)) +
  geom_error_pointrange(aes(error_neg = uptake.s, error_pos = NA),
                        size = 0.6, linewidth = 1,
                        linetype_neg = "solid",
                        colour = "firebrick") +
  facet_wrap(~ Treatment) +
  labs(title = "pointrange one-sided")
```

> Passing `0` instead of `NA` still works for backward compatibility but is soft-deprecated and will warn.

## Extending ggerror

If you know of, or need, a new error geom, please open an issue on the [GitHub repository](https://github.com/iamyannc/ggerror/issues). My first motiviation was to simplify the heck out of the error geoms, reducing the aesthetics to a single `error` aesthetic. The rest (asymmetric, one-sided, etc) are just niceties that I added along the way.

```{=html}
<!-- ============= NOTE ============
Review against `dev/new-vignettes.txt`:

- This vignette is coherent, but it still reads more like an API tour than a
  use-case document. If that is intentional, keep it as the "geom API" vignette
  and let `use-cases.Rmd` carry the modelling narrative. If not, compress the
  exposition and make each section answer a concrete question.

- The "Simple and defaults" section is the strongest part: it proves the default
  behaviour, then shows pinned wrapper vs `error_geom` cleanly. I would keep this
  structure.

- The asymmetric section needs a different example. `error_neg = sd` and
  `error_pos = sqrt(var)` are mathematically the same, so the plot does not
  actually show asymmetry. Replace with a pair that produces visibly different
  sides (`sd` vs `se`, `se` vs `ci`, quantile distances, etc.).

- The one-sided section currently gives 3 variants. That is useful for drafting,
  but probably 1 main plot plus 1 optional alternative is enough for the final
  vignette. Otherwise the section starts to feel like a gallery instead of a
  guided story.

- The plan mentions showing actual code in titles / annotations. You are already
  doing some of that with `labs(title = ...)`; if you continue, be selective.
  Too much inline code in titles will make these plots feel like screenshots of
  syntax rather than explanations.

- Minor portability point: `family = "consolas"` is Windows-specific enough that
  I would either drop it or switch to a generic `family = "mono"` if you want a
  code-like title face in rendered vignettes.
============ NOTE ============ -->
```
