---
title: "Using ggerror"
output:
  rmarkdown::html_vignette:
    css: vignette.css
vignette: >
  %\VignetteIndexEntry{Using ggerror}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 4,
  dpi = 96,
  warning = FALSE,
  message = FALSE
)
```

```{r viz-setup, include=FALSE}
library(ggplot2)
library(ggerror)

set_theme(
  theme_minimal(base_size = 13) +
    theme( 
      plot.title = element_text(
        family = "mono",
        size = 11,
        hjust = 0),
      plot.caption = element_text(face = 'italic'),
      legend.position = 'none'
      )
  )
```

## Why `ggerror`?

`ggplot2` ships several error geoms. `ggerror` started as a wrapper to simplify their usage, by dropping `x/y`-`min/max` with a one aesthetics **`error`**. This doesn't just simplify the API, but it also shift the focus to the actual intention: to visualize some magnitude as error that extends from a specified value (or a central metric, more on that at `vignette("use-cases")`).

`ggerror` is no longer just a wrapper, but it offers extra functionality compared to core `ggplot2`'s error geoms.

This vignette demonstrates the basic usage of the package. For `stat_error()` summaries and `sign_aware` erros, see `vignette("use-cases")`.

## Setup

We use `airquality` — Daily air quality measurements in New York, May to September 1973.

This dataset has 153 rows with 6 columns: 4 continuous measurements, Month and Day.

```{r data-prep}
data("airquality"); airq <- airquality

# It wouldn't be an R workflow without minimal data cleaning...
day_in_month <- function(day_in_month, month, year) {
  days_abbr <- format(as.Date(sprintf("%d-%02d-%02d", year, month, day_in_month)), "%a")
  factor(days_abbr, levels = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun"), ordered = TRUE)

}
airq$Day <- day_in_month(airq$Day, airq$Month, 1973)
airq$Month <- factor(airq$Month, labels = month.abb[5:9])


# summary table, grouped by month with Temp's means and standard deviation 
aq_monthly_avg <- data.frame(
  Month = unique(airq$Month),
  Temp = tapply(airq$Temp, airq$Month, mean),
  sd_temp_monthly   = tapply(airq$Temp, airq$Month, sd)
)

p <- ggplot(aq_monthly_avg, aes(Temp, Month))
```

## Simple and defaults

A single `error` aesthetic is enough — `ggerror` infers orientation from the data (here: numeric x axis, discrete y axis) and picks `geom_errorbar` as the default base geom.

```{r simple-default}

p + 
  geom_error(aes(error = sd_temp_monthly),
             error_geom = "pointrange",
             size = 0.7, shape = 18) +
  labs(title = "p + geom_error(aes(error = sd_temp_monthly)",
       caption = "p <- ggplot(aq_monthly_avg, aes(Temp, Month))"
       )
```

Same data, pinned `geom_error_pointrange()` wrapper:

```{r simple-pinned}
p +
  geom_error_pointrange(aes(error = sd_temp_monthly),
                        size = 0.8,
                        shape = "x") +
  labs(title = "p + geom_error_pointrange(aes(error = sd_temp_monthly))")
```

Same data, swapped via the `error_geom` argument:

```{r simple-arg}
p +
  geom_error(aes(error = sd_temp_monthly),
             error_geom = "linerange") +
  labs(title = "geom_error(error_geom = \"linerange\")")
```

::: {.alert .alert-info}
💡 **Tip:** Because `error_geom` is a regular argument, you can iterate over it functionally:
:::

```{r fp-tip, eval = FALSE}
purrr::map(c("errorbar", "linerange", "crossbar", "pointrange"),
           ~ p + geom_error(aes(error = sd_temp_monthly), error_geom = .x))
```

## Asymmetric errors

`error_neg` and `error_pos` extend in opposite directions regardless of orientation. They're useful when the dispersion measure on each side carries different meaning — say, **standard deviation** on the low side and **standard deviation × 1.5** on the high side, or here, **SD** vs **√Var** as a sanity check that they're equal.

```{r asym-prep-plot}
may_week <- subset(airq[1:7,], Month == 'May')

may_summary <- data.frame(
  Day = may_week$Day,
  Temp = may_week$Temp,
  dist2min = may_week$Temp - min(may_week$Temp, na.rm = TRUE),
  dist2max = max(may_week$Temp, na.rm = TRUE) - may_week$Temp
) # each Temp point now has a value of it's distance from the min Temp in the dataset, and a value of the distance from the max Temp in the data set.

```

```{r asym-sd-vs-var}

ggplot(may_summary,aes(x = Temp, y = Day)) +
  geom_error(aes(error_neg =  dist2min,
                 error_pos = dist2max),
             error_geom = "pointrange",
             colour_neg = "steelblue",
             colour_pos = "firebrick",
             linewidth  = 1
             ) +
  labs(title = "geom_error(error_neg = dist2min, error_pos = dist2max")
```

Deciding what consists of the negative or the positive error is up to your discretion.

Per-side styling extends to `colour`, `fill`, `linewidth`, `linetype`, `alpha`, and `width`. Pass either the side-specific form (`colour_neg`, `width_pos`) or the shared form (`colour`, `width`);

```{r asym-crossbar}
ggplot(may_summary, aes(Temp, Day)) +
  geom_error(aes(error_neg = dist2min,
                 error_pos = dist2max),
             error_geom = "crossbar",
             fill_neg = "#d97757", fill_pos = "#ffaa66",
             width_neg = 0.3,     width_pos = 0.6) +
  labs(title = "crossbar with per-side fills + widths")
```

## One-sided bars

Sometimes we care mostly about errors torwards one side, and dismis the other as uninteresting. For example, how each observation falls *below* a threshold. Set the unused side to `NA` and `ggerror` suppresses the cap and stem on that side automatically.

```{r one-sided-basic}
ggplot(may_summary, aes(Temp, Day)) +
  geom_error(aes(error_neg = dist2min, error_pos = NA),
             color = "steelblue",
             linewidth = 1,
             linetype = 9) +
  geom_point(size = 1.5) +
  labs(title = "geom_error(aes(error_neg=dist2min, error_pos = NA))")
```

**Remember:** The errors are simply a magnitude. They dont even have to be treated as errors, like in the last couple examples, where they are treated as distances from some arbitrary value (In this one-side example: the distance of each Temperature from the minimum temperature that was measured).

> Passing `0` instead of `NA` works but is discouraged, for several reasons: 1. `NA` is more explicit than `0`. 2. Since `0` is treated as a value, it won't automatically cap out the undesired side of the error bars, for example.

**See `vignette("use-cases")` for more advanced options**.

## Extending ggerror

If you know of, or need, a new error geom, please open an issue on the [GitHub repository](https://github.com/iamyannc/ggerror/issues). My first motivation was to simplify the heck out of the error geoms, reducing the aesthetics to a single `error` aesthetic. The rest (asymmetric, one-sided, etc) are just niceties that I added along the way.
