---
title: "Using ggerror"
output:
  rmarkdown::html_vignette:
    css: vignette.css
vignette: >
  %\VignetteIndexEntry{Using ggerror}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 4,
  dpi = 96,
  warning = FALSE,
  message = FALSE
)
```

## Why `ggerror`?

Base `ggplot2` ships several error geoms, and using them means
deciding on orientation, which axis the error follows, and how to
wire the lower and upper bounds correctly. That overhead piles up
fast. `ggerror` collapses it into one `error`-focused API: one
aesthetic for symmetric, two for asymmetric, and `NA` for one-sided.

This vignette walks the geom API on a single dataset. For
`stat_error()` summaries and `sign_aware` residuals, see
`vignette("use-cases")`.

## Setup

We use `CO2` — CO₂ uptake by *Echinochloa crus-galli* plants from
Quebec and Mississippi, chilled overnight or not. A small
pre-summary gives us per-group mean, standard deviation, and
variance to play with.

```{r imports}
library(ggplot2)
library(ggerror)
data(CO2)

set_theme(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(family = "consolas", size = 11, hjust = 0),
      axis.title = element_blank()
    )
)

co2_sum <- aggregate(uptake ~ Type + Treatment, data = CO2,
                     FUN = function(x) c(m = mean(x), s = sd(x), v = var(x)))
co2_sum <- do.call(data.frame, co2_sum)
co2_sum
```

## Simple and defaults

A single `error` aesthetic is enough — `ggerror` infers orientation
from the data (here: numeric `uptake.m` along x, discrete `Type`
along y) and picks `geom_errorbar` as the default base geom.

```{r simple-default}
p <- ggplot(co2_sum, aes(uptake.m, Type, colour = Treatment)) +
  geom_point(size = 3, position = position_dodge(0.4))

p +
  geom_error(aes(error = uptake.s),
             position = position_dodge(0.4)) +
  labs(title = "geom_error(aes(error = uptake.s))")
```

Same data, pinned `geom_error_pointrange()` wrapper:

```{r simple-pinned}
ggplot(co2_sum, aes(uptake.m, Type, colour = Treatment)) +
  geom_error_pointrange(aes(error = uptake.s),
                        position = position_dodge(0.4),
                        size = 0.6) +
  labs(title = "geom_error_pointrange(aes(error = uptake.s))")
```

Same data, swapped via the `error_geom` argument:

```{r simple-arg}
p +
  geom_error(aes(error = uptake.s),
             error_geom = "linerange",
             position = position_dodge(0.4),
             linewidth = 1) +
  labs(title = "geom_error(error_geom = \"linerange\")")
```

<div class="tip">
💡 **Tip:** Because `error_geom` is a regular argument, you can iterate
over it functionally — useful when scanning a few presentations to
pick one:

```{r fp-tip, eval = FALSE}
purrr::map(c("errorbar", "linerange", "crossbar", "pointrange"),
           ~ p + geom_error(aes(error = uptake.s), error_geom = .x))
```

</div>

## Asymmetric errors

`error_neg` and `error_pos` extend in opposite directions regardless
of orientation. They're useful when the dispersion measure on each
side carries different meaning — say, **standard deviation** on the
low side and **standard deviation × 1.5** on the high side, or here,
**SD** vs **√Var** as a sanity check that they're equal.

```{r asym-sd-vs-var}
ggplot(co2_sum, aes(uptake.m, Type)) +
  geom_point(size = 3) +
  geom_error(aes(error_neg = uptake.s,
                 error_pos = sqrt(uptake.v)),
             colour_neg = "steelblue",
             colour_pos = "firebrick",
             linewidth  = 1) +
  facet_wrap(~ Treatment) +
  labs(title = "error_neg = sd ; error_pos = sqrt(var) — by Treatment")
```

The same pattern works for **standard error vs confidence interval**,
**IQR halves**, or any other paired dispersion measures — swap the
two `error_*` expressions, the rest of the call stays the same.

Per-side styling extends to `colour`, `fill`, `linewidth`, `linetype`,
`alpha`, and `width`. Pass either the side-specific form (`colour_neg`,
`width_pos`) or the shared form (`colour`, `width`); side-specific
wins.

```{r asym-crossbar}
ggplot(co2_sum, aes(uptake.m, Type)) +
  geom_error(aes(error_neg = uptake.s,
                 error_pos = sqrt(uptake.v)),
             error_geom = "crossbar",
             fill_neg = "grey85", fill_pos = "grey45",
             width_neg = 0.3,     width_pos = 0.6) +
  facet_wrap(~ Treatment) +
  labs(title = "crossbar with per-side fills + widths")
```

## One-sided bars

Suppose downstream stakeholders only care about **shortfall** —
how far each group falls *below* its mean uptake — and treat
overshoot as uninteresting. Set the unused side to `NA` and
`ggerror` suppresses the cap and stem on that side automatically.

```{r one-sided-basic}
ggplot(co2_sum, aes(uptake.m, Type, colour = Treatment)) +
  geom_point(size = 3, position = position_dodge(0.4)) +
  geom_error(aes(error_neg = uptake.s, error_pos = NA),
             position = position_dodge(0.4),
             linewidth = 1) +
  labs(title = "downside only — error_pos = NA")
```

A linerange variant — heavier line, no caps — reads cleanly when
overlaid on raw points:

```{r one-sided-linerange}
ggplot(CO2, aes(uptake, Type)) +
  geom_jitter(height = 0.2, alpha = 0.3) +
  geom_error_linerange(data = co2_sum,
                       aes(x = uptake.m,
                           error_neg = uptake.s,
                           error_pos = NA),
                       linewidth  = 1.5,
                       colour     = "firebrick") +
  facet_wrap(~ Treatment) +
  labs(title = "shortfall over raw observations")
```

A pointrange variant marks the mean and the lower bound in one go:

```{r one-sided-pointrange}
ggplot(co2_sum, aes(uptake.m, Type)) +
  geom_error_pointrange(aes(error_neg = uptake.s, error_pos = NA),
                        size = 0.6, linewidth = 1,
                        linetype_neg = "solid",
                        colour = "firebrick") +
  facet_wrap(~ Treatment) +
  labs(title = "pointrange one-sided")
```

> Passing `0` instead of `NA` still works for backward compatibility
> but is soft-deprecated and will warn.

## Extending ggerror

If you know of, or need, a new error geom, please open an issue on the [GitHub repository](https://github.com/iamyannc/ggerror/issues).
My first motiviation was to simplify the heck out of the error geoms, reducing the aesthetics to a single `error` aesthetic. The rest (asymmetric, one-sided, etc) are just niceties that I added along the way.
