---
title: "Using ggerror"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using ggerror}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 4,
  dpi = 96,
  warning = FALSE,
  message = FALSE
)
```

```{r}
library(ggplot2)
library(ggerror)
```

## Why `ggerror`?

Base `ggplot2` gives you several different error geoms, but using them means
deciding which geom you need, which axis the error should follow, and how to
wire the lower and upper bounds correctly.
`ggerror` collapses that into one layer and one error-focused API.

Everything in this vignette uses `mtcars`, so you can copy the examples
directly.

## Symmetric errors

```{r}
p <- ggplot(mtcars, aes(mpg, rownames(mtcars))) +
  geom_point()

p + geom_error(aes(error = drat))
```

`geom_error()` infers orientation from the data.
When one axis is discrete and the other is numeric, the error expands along the
numeric axis.
By default it dispatches to `geom_errorbar()`.

## Choosing the dispatch geom

You can pick the base range geom either through `error_geom` or via one of the
pinned wrapper functions.

```{r}
p + geom_error(aes(error = drat), error_geom = "crossbar")
p + geom_error_crossbar(aes(error = drat))
```

Supported values are `"errorbar"`, `"linerange"`, `"crossbar"`, and
`"pointrange"`.
The wrapper functions pin that choice on purpose.

## Functional programming

When you want to generate several variants programmatically, `geom_error()`
works naturally with a vector of geom names.

```{r, eval = requireNamespace("purrr", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE)}
supported <- c("errorbar", "crossbar", "linerange", "pointrange")

plots <- purrr::map(
  supported,
  \(geom_name) p + geom_error(aes(error = drat), error_geom = geom_name)
)

patchwork::wrap_plots(plots, ncol = 2)
```

The example above is optional.
If `purrr` and `patchwork` are not installed, the rest of the vignette still
works without them.

## Asymmetric errors

Use `error_neg` and `error_pos` when the negative and positive extents differ.

```{r}
ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_point() +
  geom_error(aes(error_neg = drat / 2, error_pos = drat))
```

`error_neg` extends in the negative direction and `error_pos` extends in the
positive direction, regardless of whether the error is horizontal or vertical.
Both must be supplied together.
If you want symmetric behavior, use `error`.

## One-sided bars

Setting one side to `0` gives you a one-sided error bar.

```{r}
ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_point() +
  geom_error(
    aes(error_neg = 0, error_pos = drat),
    width_neg = 0
  )
```

On its own, `aes(error_neg = 0, error_pos = drat)` still draws a cap at the
observation value at the default width, because the negative side has zero
length but inherits the shared cap width.
Setting `width_neg = 0` removes that cap and gives you a genuine one-sided
bar: in the split `errorbar` path, `ggerror` draws only the outer caps, so
the cap at the shared bound is suppressed.

## Advanced: per-side styling

For finer control, the negative and positive halves can be styled separately
with fixed `_neg` and `_pos` parameters for `colour`, `fill`, `linewidth`,
`linetype`, `alpha`, and `width`.

```{r}
ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_point() +
  geom_error(
    aes(error_neg = drat / 2, error_pos = drat),
    colour_neg = "steelblue",
    colour_pos = "firebrick",
    linewidth_neg = 0.3,
    linewidth_pos = 1
  )
```

```{r}
ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_error(
    aes(error_neg = drat / 2, error_pos = drat),
    error_geom = "crossbar",
    fill_neg = "grey90",
    fill_pos = "grey60",
    width_neg = 0.2,
    width_pos = 0.8
  )
```

These per-side values are fixed parameters, not mapped aesthetics.
In other words, use `colour_neg = "steelblue"`, not
`aes(colour_neg = some_column)`.

For `geom_error_pointrange()`, the shared point keeps the base point
aesthetics while the line halves can still differ.
`shape`, `size`, and `stroke` are not split because there is still only one
central point per observation.
