---
title: "Use cases: stats and residuals"
output:
  rmarkdown::html_vignette:
    css: vignette.css
vignette: >
  %\VignetteIndexEntry{Use cases: stats and residuals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 4,
  dpi = 96,
  warning = FALSE,
  message = FALSE
)
```

```{r imports}
library(ggplot2)
library(ggerror)
data(CO2)

set_theme(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(family = "consolas", size = 11, hjust = 0),
      axis.title = element_blank()
    )
)
```

The `CO2` dataset records carbon dioxide uptake in *Echinochloa
crus-galli* plants from Quebec and Mississippi, chilled overnight or
not — 84 observations across 5 variables. We'll use it to show two
end-to-end use cases: summarising raw data with `stat_error()`, and
reading model residuals as direction-aware bars with `sign_aware`.

## Summarising raw data with `stat_error()`

`stat_error()` follows ggplot2's `fun.data` contract — pass a summary
function and it computes `y`, `ymin`, `ymax` per group. The default
is `mean_se` (mean ± one standard error):

```{r stat-default}
ggplot(CO2, aes(Treatment, uptake)) +
  stat_error() +
  labs(title = "stat_error()  # default = mean_se")
```

Switch to a 95% confidence interval and a `pointrange` base geom:

```{r stat-ci}
ggplot(CO2, aes(Treatment, uptake)) +
  stat_error(fun = "mean_ci", error_geom = "pointrange") +
  labs(title = 'stat_error(fun = "mean_ci", error_geom = "pointrange")')
```

The `conf.int` argument controls the CI width — `0.95` by default,
but anything in `(0, 1)` works:

```{r stat-conf}
ggplot(CO2, aes(Treatment, uptake)) +
  stat_error(fun = "mean_ci", conf.int = 0.99,
             error_geom = "pointrange") +
  labs(title = 'stat_error(fun = "mean_ci", conf.int = 0.99)')
```

### Custom summary function: median ± mean absolute deviation

The `fun` slot accepts any function that takes a numeric vector and
returns a one-row data frame with `y`, `ymin`, `ymax`. Here's a
robust dispersion summary — **median** as the centre,
**mean absolute deviation from the median** as the spread:

```{r stat-custom}
mae_summary <- function(y) {
  md  <- median(y)
  mae <- mean(abs(y - md))
  data.frame(y = md, ymin = md - mae, ymax = md + mae)
}

ggplot(CO2, aes(Type, uptake)) +
  stat_error(fun = mae_summary, error_geom = "crossbar",
             fill = "grey90") +
  facet_wrap(~ Treatment) +
  labs(title = "stat_error(fun = median ± mean(|x − median|))")
```

Custom functions can take extra parameters too — `stat_error()`
forwards any matching named arguments via its signature-aware
dispatcher. (For one-off knobs, just close over them; for plumbed
arguments, name them in the function and pass at the call site.)

> **Aside** — `geom_error(stat = "error", fun = ...)` is the
> layer-level equivalent of `stat_error()`. Useful when you want
> the geom call to lead and the stat to be incidental.

## Residual diagnostics with `sign_aware`

`sign_aware = TRUE` interprets the sign of `error` as direction:
positive values extend in the positive axis direction, negative in
the negative. That single switch turns residuals into a self-coloring
diagnostic layer.

Fit a model and attach predictions and residuals:

```{r fit-model}
model   <- lm(uptake ~ conc, data = CO2)
co2_fit <- transform(CO2,
                     predicted = predict(model),
                     residual  = resid(model))
```

Now plot the fitted line, the actual observations, and the residuals
all in one layer. Each bar runs from the fitted curve up or down to
the corresponding observation. Colour says direction:

```{r residuals-plot, fig.height = 4.5}
ggplot(co2_fit, aes(conc, predicted)) +
  geom_line(linewidth = 0.4, colour = "grey40") +
  geom_point(aes(y = uptake), alpha = 0.5) +
  geom_error(aes(error = residual),
             sign_aware  = TRUE,
             orientation = "x",
             colour_pos  = "firebrick",
             colour_neg  = "steelblue",
             linewidth   = 0.5) +
  labs(title = paste(
    "lm(uptake ~ conc) |> resid()",
    "geom_error(error = residual, sign_aware = TRUE)",
    sep = "\n"
  ))
```

**Reading it**: each red bar marks a plant whose actual uptake
exceeded the model's prediction (under-prediction); each blue bar
marks a plant the model over-predicted. Bar **length** is the
residual magnitude. The clustering of same-coloured runs along
`conc` is informative on its own — a hint of unmodeled structure.

Because both axes are numeric here, `ggerror` can't infer the
direction the bar should extend, so we pass `orientation = "x"`
explicitly. (When one axis is discrete, orientation is inferred.)

A few diagnostic notes:

- A **zero residual** produces no bar — there's nothing to draw.
- An **`NA` residual** also draws nothing, and triggers a row-indexed
  warning so you can find the offending observation.
- `sign_aware = TRUE` is incompatible with `stat = "error"` — the
  stat already produces symmetric bounds. The combination errors at
  layer construction.
