---
title: "Use cases: stats and residuals"
output:
  rmarkdown::html_vignette:
    css: vignette.css
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Use cases: stats and residuals}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 4,
  dpi = 96,
  warning = FALSE,
  message = FALSE
)
```

```{r viz-setup, include=FALSE}
library(ggplot2)
library(ggerror)

set_theme(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(
        family = "mono",
        size = 11,
        hjust = 0),
      plot.caption = element_text(face = 'italic'),
      legend.position = 'none'
      )
  )
```

This vignette demonstrates more advanced usage of the package. For a
"get-started" and overview of this package's basic usecase, see
`vignette("ggerror")`.

## Setup

We use `airquality` — Daily air quality measurements in New York, May to
September 1973.

This dataset has 153 rows with 6 columns: 4 continuous measurements,
Month and Day.

```{r data-prep}

# It wouldn't be an R workflow without minimal data cleaning...
day_in_month <- function(day_in_month, month, year) {
  days_abbr <- format(as.Date(sprintf("%d-%02d-%02d", year, month, day_in_month)), "%a")
  factor(days_abbr, levels = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun"), ordered = TRUE)
}
airq$Day <- day_in_month(airq$Day, airq$Month, 1973)
airq$Month <- factor(airq$Month, labels = month.abb[5:9])

```

\> NOTE FROM HUMAN TO LLM: \>

Style improvement: Beneath ##SETUP, add callout-info class that is
collapsible, and collapsed by default, with the data info and more from
`airquality` docs. Offer the users to expand the following info box for
more information, or advise them to just type `?airquality` for more
information. vignette should support collapsible HTML (it is rendered to
a pkgdown website, im not mistaken - verify!).

ALSO: for the summary function, please add valid LaTeX to show the math
definition of these functions. IF YOU STRUGGLE WITH LATEX: leave a
placeholder for an image, i will take a screenshot. but prioritise to
render this nitly with latex.

\> \> END NOTE

Let's use it to show two use cases: Plot sign-aware residual plot and
summarizing raw data with `stat_error()`.

## Residual plot with `sign_aware`

`sign_aware = TRUE` interprets the sign of `error` as direction:
positive values extend in the positive axis direction, negative in the
negative.

#### Let's fit a linear model and see how `Temp` can be explained by `Wind`.

```{r fit-model}
model   <- lm(Temp ~ Wind, data = airq)
airq_fit <- cbind(airquality[,c('Temp','Wind')],
                  predicted = predict(model),
                  residuals = resid(model)
                 )
```

Let us compliment usual ggplot2's workflow for plotting residual vs true
vales:

```{r residuals-plot, fig.height = 4.5}
ggplot(airq_fit, aes(x = Wind, y = predicted)) +
  geom_line(linewidth = 0.4, colour = "grey40") +
  geom_point(aes(y = Temp), alpha = 0.5) +
  geom_error(aes(error = residuals),
             sign_aware  = TRUE,
             orientation = "x",
             color_pos  = "firebrick",
             color_neg  = "steelblue",
             linewidth   = 0.5,
             alpha       = 0.85) +
  labs(title = paste(
    "[1] residuals = resid(lm(Temp ~ Wind))",
    "[2] geom_error(error = residuals, sign_aware = TRUE)",
    sep = "\n"
  ))
```

- **Flexible styling**: `ggerror` provides us with per-direction styling
  (`color`) as well as global styles (`alpha`,`linewidth` etc..). How to
  interpret the plot is up to you. You may want to color the direction
  differently, to show not the error's sign, but the direction to which
  it extends. Experiment this by tweaking either the actual color value,
  or the `neg/pos` suffix.

- **`orientation = "x"`** Because both axes are numeric, `ggerror` can't
  infer the direction the bar should extend, so we pass
  `orientation = "x"` (which extends the bar along the y axis)
  explicitly. When one axis is discrete, orientation is inferred.

- **NA values vs 0:** An **`NA` value** draws nothing, and triggers a
  row-indexed warning. As in `ggplot2`'s ecosystem, we would usually
  exclude invalid (missing or infinite) values before plotting. 0 is
  treated as any other value, and "draws" a bar of length zero
  (basically, nothing). using `NA` comes in handy when we want to
  explicitly draw one-sided bars (see `vignette("ggerror")`).

## Summarizing raw data with `stat_error()`

`stat_error()` follows ggplot2's `fun.data` - pass a summary function
and it computes `y`, `ymin`, `ymax` per group. Regardless of the
orientation of your plot, the summary function **must return a data
frame with these exact column names**.

The default summary function is `mean_se` (mean ± one standard error):

```{r stat-default}
ggplot(airq, aes(Month, Temp)) +
  stat_error(width = 0.2) +
  stat_summary(geom='point') +
  labs(title = "stat_error() with default = mean_se") 
```

I equipped `ggerror` with two summary functions. The second one is
`mean_ci` which computes the mean and 95% confidence interval error
bars.

```{r stat-ci}
ggplot(airq, aes(Month, Temp)) +
  stat_error(fun = "mean_ci", error_geom = "pointrange") +
  labs(title = 'stat_error(fun = "mean_ci", error_geom = "pointrange")')
```

The `conf.int` argument controls the CI width — `0.95` by default, but
valid probability *(0-1)* works:

```{r stat-conf}
ggplot(airq, aes(Month, Temp)) +
  stat_error(fun = "mean_ci", conf.int = 0.99,
             error_geom = "linerange") + 
  stat_summary(geom="point") +
  labs(title = 'stat_error(fun = "mean_ci", conf.int = 0.99)')
```

> **Note:** I combine multiple summary geoms. `ggerror` is compatible
> with `ggplot2` (core) and it's ecosystem.

### Custom summary function: median ± mean absolute deviation

You shall not be limited by my imagination. Create your own custom
summary functions, as long as they obey the rules I mentioned earlier
(returned object must be a data frame with three columns: `y`, `ymin`,
`ymax` and one numeric row). R's `mad()` computes the median absolute
deviation from the median. Let us use the **mean** absolute deviation
(from the median), and call it `mae` (`e` for error, to avoid naming
conflicts).

**median** as the center, **mean absolute deviation from the median** as
the spread:

```{r stat-custom}
mae_summary <- function(my_vec, scale_by = 1) {
  md  <- median(my_vec)
  mae <- mean(abs(my_vec - md)) * scale_by
  data.frame(y = md, ymin = md - mae, ymax = md + mae)
}

ggplot(airq, aes(Month, Temp)) +
  stat_error(fun = mae_summary, error_geom = "crossbar",
             fill = "grey90") +
  labs(title = "stat_error(fun = median ± mean(|x − median|))")
```

Custom functions can take extra parameters too — `stat_error()` forwards
any matching named arguments in the same way you are used to use the
`...` argument. In this example, in order to multiply the error by 2,
you'd write `stat_error(fun = mae_summary, scale_by = 2)`

\> NOTE FROM HUMAN TO LLM \> If possible, add another collipsable 
\> HUMAN NOTE \>
I'd like to add add {r aditional-code-collapsed} as additional collapsable chunck. To avoid explosion of plots, It should be collappsed by default.

Another important question: In both vignettes, I am using this syntax for reference: `vignette("vignette-name")`. Will it actually be clickable (in RStudio's help window and/or my website)?

```{r aditional-code-collapsed}

ggplot(airq, aes(Month, Wind)) +
  stat_error(fun = mae_summary,scale_by = 2, error_geom = "crossbar") +
  stat_summary(geom = "point")

```

\> END NOTE\>

> **Aside** — `geom_error(stat = "error", fun = ...)` is the layer-level
> equivalent of `stat_error()`, which by default uses
> `stat = "identity"`.
