---
title: "stat_error() and sign_aware + demo"
output:
  rmarkdown::html_vignette:
    css: vignette.css
vignette: >
  %\VignetteIndexEntry{stat_error() and sign_aware + demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 4,
  dpi = 96,
  warning = FALSE,
  message = FALSE
)
```

```{r imports-ggplot2-settings}
library(ggplot2)
library(ggerror)

set_theme(
  theme_minimal() + theme(
  plot.title = element_text(hjust = 0, family = "consolas", size = 12),
  axis.title = element_blank()
))

```

Version 1.0.0 adds two new features:

- **`stat_error()`** computes error bounds directly from raw data. It follows ggplot2's `fun.data` contract, meaning it works
  with any summary function you pass in - `mean_se`, `mean_ci`, or a custom one *(see details below)*.
- **`sign_aware = TRUE`** interprets the sign of input values as direction. Positive and negative values
  are mapped to `error_pos` and `error_neg`, respectively.
  This is useful when plotting residuals or any signed quantity where direction carries meaning.

## `stat_error()`:

Pass raw observations and let the stat compute the error. The default
is `mean_se` - mean with one standard error:

```{r stat-default}
ggplot(mtcars, aes(factor(cyl), mpg)) +
  stat_error()
```

`mean_ci` for a 95% confidence interval:

```{r stat-ci}
ggplot(mtcars, aes(factor(cyl), mpg)) +
  stat_error(fun = "mean_ci", error_geom = "pointrange")
```

*Both `mean_ci` and `mean_se` accept `na.rm` as an argument.*
*`mean_ci` also accepts `conf.int`, default is 0.95.*

### Custom summary function
You can also pass a custom function following ggplot2's `fun.data`
syntax — it takes a numeric vector and returns a single-row
data frame with columns `y`, `ymin`, `ymax`:

```{r stat-custom}
iqr <- function(y, type = 6) {
  data.frame(
    y    = median(y),
    ymin = stats::quantile(y, 0.25, names = FALSE, type = type),
    ymax = stats::quantile(y, 0.75, names = FALSE, type = type)
  )
}

ggplot(mtcars, aes(factor(cyl), mpg)) +
  stat_error(fun = iqr, type = 1)
```

You can pass an arbitrary number of arguments to `stat_error()`.

While `geom_error` uses `stat = 'identity'`, you could also pass `stat = 'error'`, which is equivalent to `stat_error()`.

## `sign_aware`: residuals as one-sided bars
I really love this usecase. Drawing sign-aware errors (with colours!) used to be a pain. Not anymore!

- **First, let us fit a (dummy) model**
```{r fit-model}
model <- lm(mpg ~ wt, data = mtcars)
mt_model <- mtcars
mt_model$predicted <- fitted(model) # predicted values (y_hat)
mt_model$res <- resid(model) # raw distances (y-y_hat), some positive, some negative
```

```{r residuals-plot}
ggplot(mt_model, aes(mpg,predicted)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  geom_error(aes(error = res), sign_aware = TRUE, orientation = "x",
             colour_pos = "firebrick", colour_neg = "steelblue")
```

*Since both axes are numeric, `ggerror` can't infer the orientation itself, so we have to pass it